---
title: "simple-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simple-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# install packages

```{r eval=FALSE}
install.packages("tidyverse")
install.packages("snakecase")
install.packages("h2o")
install.packages("remotes")
remotes::install_github('ML4LHS/gpmodels')
```

# setup

```{r setup}
library(tidyverse)
library(snakecase)
library(gpmodels)
library(lubridate)
```

https://data.world/informatics-edu/synthea-synthetic-patient-information

https://synthea.mitre.org/downloads

# Import data

```{r}
temp <- tempfile()
download.file("https://synthetichealth.github.io/synthea-sample-data/downloads/synthea_sample_data_csv_apr2020.zip", temp)
unzip(temp)
```

# Format data

```{r}
patients = read_csv("./csv/patients.csv") %>% 
  janitor::clean_names() %>% 
  # mutate(description = to_snake_case(description)) %>% 
  select(patient = id, birthdate, deathdate, marital, race, ethnicity, gender)

patients %>% str()

conditions = read_csv("./csv/conditions.csv") %>% 
  janitor::clean_names() %>% 
  mutate(description = to_snake_case(description), category = "condition", value = 1) %>% 
  select(patient, encounter, time = start, category, description = description)

conditions %>% str()

observations = read_csv("./csv/observations.csv") %>% 
  janitor::clean_names() %>% 
  unite(col = "description", c("description", "units")) %>% 
  mutate(description = to_snake_case(description), category = "observation") %>% 
  select(patient, encounter, time = date, category, description, value)

observations %>% str()

encounters = read_csv("./csv/encounters.csv") %>% 
  janitor::clean_names() %>% 
  filter(encounterclass == "inpatient") %>% 
  mutate(description = to_snake_case(description)) %>% 
  select(patient, encounter = id, start, stop, encounterclass, encounter_descr = description)
```

# Define static outcome

> A static outcome must be defined and joined with the fixed data

```{r}
outcome_pne = conditions %>% 
  filter(grepl("pneumonia", description)) %>% 
  group_by(encounter) %>% 
  slice_min(time) %>% 
  ungroup() %>% 
  select(patient, encounter, pneumonia_time = time)

outcome_pne %>% str()
```

# Build cohort and demographic data (fixed data)

```{r}
fixed_data = patients %>% 
  inner_join(encounters %>% filter(encounterclass == "inpatient") %>% unique() %>% arrange(desc(start))) %>% 
  left_join(outcome_pne %>% select(patient, pneumonia_time)) %>% 
  mutate(pneumonia = if_else(!is.na(pneumonia_time), "PNA", "No_PNA") %>% factor(levels = c("PNA", "No_PNA"))) %>% 
  mutate(pneumonia = if_else(!is.na(pneumonia_time), 1, 0, 0) %>% as.factor()) %>% 
  mutate(pred_start = start, pred_stop = coalesce(deathdate, pneumonia_time, stop)) %>% 
  select(-c(start, stop)) %>% 
  filter(!is.na(pred_start), # prediction start must be present
         pred_start <= pred_stop) %>%  # prediction start should never be greater than prediction stop
  distinct(across(c(encounter)), .keep_all = TRUE)
```

# Temporal data

```{r}
temporal_data = bind_rows(conditions, observations) %>% 
  select(-encounter) %>% 
  inner_join(fixed_data %>% select(patient, encounter)) %>% 
  mutate(time = as_datetime(time))
```

# Define data model

> start predictions: admission
> end predictions: discharge
> prediction interval: 6 hours

```{r}
tf = gpmodels::time_frame(fixed_data = fixed_data,
                          temporal_data = temporal_data, # timestamped data
                          fixed_id = 'encounter', 
                          fixed_start = 'pred_start', 
                          fixed_end = 'pred_start', 
                          temporal_id = 'encounter', 
                          temporal_time = 'time', 
                          temporal_variable = 'description', 
                          temporal_value = ,'value',
                          temporal_category = 'category', 
                          step = hours(6), # how often to make predictions
                          max_length = days(7), # maximum time limit to make predictions (e.g. 7 days inpatient)
                          output_folder = './gpmodels_out/',
                          create_folder = TRUE,
                          chunk_size = 2000)
```

```{r}
tf$fixed_data %>% 
  janitor::tabyl(pneumonia)
```

```{r}
tf = tf %>% 
  pre_dummy_code()
```

# Define custom functions

```{r}
presence = function (x) { if_else(length(x) > 0, 1, 0, 0) }
```

# Define predictors

```{r}
future::plan('multisession')

tf %>% add_baseline_predictors(category = "condition", lookback = months(12), window = months(3), stats = c(presence = presence))

tf %>% add_baseline_predictors(category = "observation", lookback = months(12), window = months(3), stats = c(min = min, max = max, median = median, last = dplyr::last))
```


# Define rolling outcome

> pneumonia in the next 24 hours

```{r}
tf %>% add_rolling_outcomes(variables = "pneumonia", lookahead = hours(24), stats = c(presence = presence))
```

# Randomly split patients into groups for cross-validation

```{r}
set.seed(1234)
folds = tf$fixed_data %>%
  select(patient) %>%
  distinct() %>%
  group_by(patient) %>%
  mutate(fold = sample.int(6, patient %>% length(), replace = TRUE))

folds %>% write_csv(file.path("folds.csv"))

folds %>% janitor::tabyl(fold)
```

# Combine prepped data

```{r}
gpm_prepped = tf %>% 
  combine_output() %>% 
  write_csv("prepped_data.csv")
```

# Build model

```{r}
response = "pneumonia"
predictors = setdiff(names(gpm_prepped), c(response,"outcome_pneumonia_presence_24", gpm_prepped %>% select(patient:pred_stop) %>% names()))
predictors %>% length()
# predictors

library(h2o)
h2o.init(nthreads=-1) # multiprocessing support
h2o.removeAll() # Clean slate - just in case the cluster was already running
```

# Prep modeling datasets

```{r}
training = gpm_prepped %>% inner_join(folds %>% filter(fold %in% 1:4)) %>% 
  mutate(pneumonia = pneumonia %>% as.factor())

validation = gpm_prepped %>% inner_join(folds %>% filter(fold == 5)) %>% 
  mutate(pneumonia = pneumonia %>% as.factor()) %>% select(-fold)

test = gpm_prepped %>% inner_join(folds %>% filter(fold == 6)) %>% 
  mutate(pneumonia = pneumonia %>% as.factor()) %>% select(-fold)
```

# Build and evaluate model

```{r}
gbm_pna_static = h2o.gbm(x = predictors,
                         y = response,
                         training_frame = training %>% as.h2o(),
                         fold_column = "fold",
                         validation_frame = validation %>% as.h2o(),
                         ntrees = 200,
                         learn_rate = 0.05,
                         learn_rate_annealing = 0.99,
                         score_tree_interval = 10,
                         stopping_rounds = 5,
                         stopping_tolerance = 1e-5,
                         stopping_metric = "AUC",
                         seed = 1234,
                         keep_cross_validation_predictions = TRUE)

gbm_pna_static %>%
  h2o.performance(newdata = test %>% as.h2o()) %>%
  h2o.auc() %>%
  round(3)
```

```{r}
gbm_pna_static_importance = h2o.varimp(gbm_pna_static) %>%
  arrange(desc(relative_importance)) %>% 
  rename(pc_relative_importance = relative_importance, pc_scaled_importance = scaled_importance, pc_percentage = percentage)

gbm_pna_static_importance %>%
  as.data.frame() %>%
  head(20) %>%
  mutate(pc_relative_importance = pc_relative_importance %>% round(2)) %>%
  select(variable, pc_relative_importance) %>%
  arrange(desc(pc_relative_importance)) %>%
  ggplot(aes(x = reorder(variable, pc_relative_importance), y = pc_relative_importance)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  xlab("")
```

# References

Jason Walonoski, Mark Kramer, Joseph Nichols, Andre Quina, Chris Moesel, Dylan Hall, Carlton Duffett, Kudakwashe Dube, Thomas Gallagher, Scott McLachlan, Synthea: An approach, method, and software mechanism for generating synthetic patients and the synthetic electronic health care record, Journal of the American Medical Informatics Association, Volume 25, Issue 3, March 2018, Pages 230–238, https://doi.org/10.1093/jamia/ocx079
